HealpixNet: Efficient spherical Convolutional Neural Network with HEALPix sampling for cosmological applications

### 2018-08-28 scnn.tex
changes from last week

% \documentclass[preprint,12pt,authoryear]{elsarticle}
\documentclass[final,twocolumn,3p,times,authoryear]{elsarticle}

%% Use the option review to obtain double line spacing
%% \documentclass[authoryear,preprint,review,12pt]{elsarticle}

@@

%\usepackage{lmodern}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}

@@

%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
\usepackage{color}
\usepackage{url}
% \usepackage{hyperref}
\usepackage[draft]{hyperref}
\usepackage{graphicx,array} \graphicspath{{figures/}}
\usepackage{subcaption}
\usepackage{amsmath}
\usepackage{siunitx}
\usepackage{bm}
\usepackage{aasmacros}

\usepackage{enumitem} % to remove when done with todo list

%% The lineno packages adds line numbers. Start line numbering with
%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
%% for the whole article with \linenumbers.
%% \usepackage{lineno}

@@

\author[SDSC]{Nathanaël Perraudin}
\author[EPFL]{Michaël Defferrard}
\author[ETHZ]{Tomasz Kacprzak}

@@

Convolutional Neural Networks (CNNs) are now a cornerstone of the Deep Learning toolbox and have led to many breakthroughs in Artificial Intelligence.
So far, these networks have mostly been developed for regular Euclidean domains such as those supporting images, audio, or video.
Because of their success, CNN-based methods are becoming increasingly popular in Cosmology.
Cosmological data often comes in a form of spherical maps, which makes the use of traditional CNNs more complicated.
The commonly used pixelzation scheme for spherical maps is HEALPix, which uses an equal-area, isolatitude sampling.
We present a spherical CNN for analysis of full and partial HEALPix maps, which we call HealpixNet.
We construct a spherical CNNs by representing the sphere as a graph.
Graphs are versatile data structures which can represent pairwise relationships between objects or act as a discrete representation of a continuous manifold.
Using the graph-based representation, we define many of the standard CNN operations, such as convolution and pooling, with filters restricted to being radial.
This way, HealpixNet is a special case of a graph CNN, tailored to the HEALPix sampling of the sphere.
% TBD if we include it
% This approach is very efficient computationally than using spherical harmonics to perform convolutions, an alternative that has been explored previously.
An advantage of this representation is that the convolution operations with this approach are very efficient, which enables the algorithm to run very fast.
We demonstrate the method on a classification problem of mass maps from two cosmological models and compare the performance of the CNN with that of two baseline classifiers, based on the power spectrum and pixel density histogram.
Our experimental result shows that the performance of HealpixNet to always be better or equal to both of these baselines.
For high noise levels and for data covering only a smaller fraction of the sphere, HealpixNet achieves typically 10\% better classification accuracy than those baselines.
Finally, we propose a visualization of filters learned by the HealpixNet from the mass maps as some analysis tool for the neural network.
Code and examples are available at \todo{\url{https://github.com/SwissDataScienceCenter/HealpixNet}}.

@@

\section*{TODO list}
\begin{itemize}[noitemsep,topsep=0pt,parsep=0pt,partopsep=0pt]
    \item (TK) check text figure 1
	\item (TK) Related work in cosmo. The only cosmological applications of spherical CNNs in the related work are about projection on the tangent plane \citep{schmelze2017cosmologicalmodel, fluri2018deep, gupta2018nongaussianinformation}. Aren't there applications in cosmology for the other (c.f. related work section) approaches? Most of them are illustrated in omnidirectional imagery. If none, say it and mention some alternatives, e.g., wavelets?
	\item (MD) distinction between CNN and FCN
	\item (MD) check paragraph on our chosen architecture in the Experiments sections
    \item (MD) check vertex interpretation in Efficient convolutions. Powers of L are weights of paths.
    \item (TK) check if papers which appear as arxiv were not published in a conference or journal
    \item (TK) check part sphere notebook and whole sphere
    \item (TK) read everything
    \item (NP) Put Raphael in the dataset author. 
    \item (done!) (TK) add small description of maps\_downsampled\_64.npz in the whole\_sphere notebook. \nati{We need to talk about that.}
	\item (TK) Example applications for dense prediction (classification and regression) for cosmology (color text in 2.8)
	\item (Probably OK) Beware to not confuse reader with synonyms, e.g. pixelization and sampling.
	\item (Should be done) Check names: we use HealpixNet, HealpixNet (CNN variant), HealpixNet (FCN variant)
    \item (I think it is OK) (NP) Check the wording. Be sure that the convolution on SO(3) is clear: we want to say (a) is the one associated with the SO(3) rotation group (b) can be performed  in the spectral domain by a multiplication (c) is equivariant to rotation (d) is described well in Section 2 of \citep{kondor2018clebsch}.
	\item (Done) (NP) Change equation after (2) and check all kroneker
	\item (Everyone should check this) Notation: write element $i$ of vector $y$ as $y_i$, $y(i)$, or $y[i]$? Be consistent. Same for matrices. Keep indices.
	\item (TK) Font: do we have to use times (set in the documentclass)? I (MD) find the rendering of math a bit ugly.
    \item (TK) fix the bibliography
    \item Check links in footnotes.

@@

% \todo{@Tomek: can you explain why we care about parts of the sphere? (that is an advantage of the graph model over the spherical harmonics)}
% \todo{@Tomek: How is the data acquired? Do telescopes and satellites take measurements according to the HEALPix sampling? (it is also an advantage of our method to be agnostic to the sampling, which might be useful to researchers working with missing data or different models)}
% \nati{@Tomek: I think what Michael has in mind is that the graph approach is very versatile and extendable. It can be used with other samplings (even irregular ones) or only part of the sphere. This flexibility might be an advantage to mention. In particular, compared to~\citet{cohen2018sphericalcnn} it allows us to do only do the required computation when only a part of the sphere is considered.}

@@

% cohen2017convolutional (ICML workshop) and cohen2018sphericalcnn (ICLR best paper) are redundant
A first direction to generalize CNNs to spherical data was to apply a standard 2D CNN to a grid discretisation of the sphere~\citep{boomsma2017spherical,su2017sphericalconv,coors2018spherenet}. While this approach uses the well-developed 2D convolution and hierarchical pooling, it is restricted to those specific grid samplings which are highly irregular on the sphere and hence inherently deform the domain where the data is supported. 
A second direction attempted by~\citet{cohen2018sphericalcnn} was to leverage the spherical Fourier transform and to perform the convolution associated to the SO(3) rotation group in the spectral domain, thanks to the convolution theorem.
Contrarily to the first direction, the resulting convolution is rotational equivariant, meaning that a rotation of the input implies the same rotation of the output.
This is an important property as translation invariance is probably one of the main reasons why classic CNNs are successful in the first place.
Nevertheless, this approach is computationally expensive even if a fast spherical Fourier transforms is used. 
%(They exist for particular samplings.)
Finally, the use of a graph to model the discretised sphere was considered by~\citet{khasanova2017graphomni} as well, towards the goal of designing a  network that is invariant to rotations.
While their architecture accommodates these invariances on the sphere and remains fairly efficient, they did not take advantage of a hierarchical pixelisation and use as pixelisation a grid discretisation of the sphere. 
Moreover, the statistical layer, which introduces the invariance, needs to be hand-crafted to capture statistics of interest.

%\todo{better identify the shortcomings: (i) part of sphere / missing data, (ii) computational efficiency, (iii) equivarience}
\nati{The part of the sphere/missing data is not really there, but I still do not know how to include it!}

To overcome those shortcomings, we introduce an efficient spherical CNN formulation which uses (i) convolutions on graphs and (ii) hierarchical pooling, while working with the HEALPix sampling.
The main idea is to represent the sphere $S^2$, a 2-dimensional manifold embedded in $\mathbb{R}^3$, as a graph of connected pixels.
The length (or weight) of the shortest path between two pixels is an approximation of the geodesic distance between them on the manifold.
The use of this approach is enabled by recent developments in graph signal processing, where fundamental tools, such as for example the Fourier transform, were defined for graphs \citep{shuman2013emerging}.
Our approach uses the graph CNN formulation introduced by \citet{defferrard2016convolutional}, which can accommodate any sampling of the sphere.
Using a $k$-nearest neighbours graph, the convolution operation costs $\mathcal{O}(N_{pix})$ operations, where $N_{pix}$ is the number of pixels.
This is the lowest possible complexity for a convolution operation without approximations (e.g., by sketching).
The flexibility of modeling the data domain with a graph allows to easily model data that spans only a part of the sphere, or data that is not uniformly sampled.
Our hierarchical pooling strategy exploits a hierarchical pixelisation of the manifold to analyse the data at multiple scales.
% Such hierarchical pixelisations of the sphere have been well studied.

@@

We give a practical demonstration of the application of our package to cosmological model discrimination using convergence maps, similar to the ones used by \citet{schmelze2017cosmologicalmodel}.
In a simplified scenario, we classify convergence maps on parts of a sphere into two cosmological models.
These models were designed to have the same power spectrum in range $\ell < 1000$.
We compare the performance of our spherical CNN to a baseline algorithm, an SVM classifier which takes pixel histograms or power spectrum densities (PSDs) of these maps as input.
The comparison is made as a function of the additive noise level and the used area of the sphere. Results show that our model is better at discriminating the maps and is more resilient to noise.

@@

A CNN is composed of the following main building blocks~\citep{lecun1998cnn}:
(i) a convolution,
(ii) a non-linearity,
(iii) a down-sampling operation,
(iv) a pooling operation, and
(v), optionally,  normalization.\footnote{\citet{ioffe2015batchnorm} have shown that batch normalization helps training. We verified this experimentally in our setting as well.}
Our architecture is depicted in~\figref{architecture} and discussed in greater details in \secref{architecture}. As operations (ii) and (v) are point-wise, they do not depend on the data domain. The pooling operation is simply a permutation invariant aggregation function which does not need to be adapted either. The convolution and down-sampling operations, however, need to be generalized from Euclidean domains to the sphere.

@@

Before doing any numerical analysis on the sphere, one first has to choose a tessellation, i.e., an exhaustive partition of the sphere into finite area elements, where the data under study is quantized.
The simplicity of the spherical form belies the intricacy of global analysis on the sphere.
There is no known point set which achieves the analogue of uniform sampling in Euclidean space and allows exact and invertible discrete spherical harmonic decompositions of arbitrary but band-limited functions.
Any existing proposition of practical schemes for the discrete treatment of such functions on the sphere introduces some systematic error dependent on the global properties of the point set.
While our method is applicable to any pixelisation of the sphere, two details depend on the chosen sampling: (i) the choice of neighbours in the construction of the graph, and (ii) the choice of parent nodes when coarsening the graph.
As HEALPix~\citep{gorski2005healpix} is the most popular sampling used in cosmology, our target application, we'll tailor the method to that particular sampling in the subsequent exposition.
%We'll note in later sections where the method should be adapted for a different sampling.

@@

HEALPix is an acronym for Hierarchical Equal Area isoLatitude Pixelization of a sphere.
This pixelisation produces a hierarchical subdivision of a spherical surface where each pixel covers the same surface area as every other pixel.
A hierarchy is desired for the data locality in the computer memory.
Equal area is advantageous because white noise generated by the signal receiver gets integrated exactly into white noise in the pixel space.
Isolatitude is essential for the implementation of a fast spherical transform.
HEALPix is the sole pixelisation scheme which satisfies those three properties.

@@

Our graph is constructed as an approximation of the continuous 2D manifold that is the surface of the sphere.
Indeed, \citet{belkin2007convergence} showed that the graph Laplacian converges to the Laplace-Beltrami when the number of pixels goes to infinity.
While our construction will not exactly respect the setting defined by \citet{belkin2007convergence}, we observe empirically strong evidence of convergence.

@@

Finally, the graph convolution can be interpreted in the vertex domain as a scalar product with localizations $\T_i h$ of the kernel $h$. Indeed, the result of the convolution of the signal $\f$ with the kernel $h$ is
\begin{equation} \label{eqn:graph_convolution_spatial}
	(h(\L) \f)_i = \langle \T_i h(\L), \f \rangle = \langle h(\L) \b \delta^i, \f \rangle.
\end{equation}
To make the parallel with the classical 1D convolution, let $f, g\in \mathbb{Z} \rightarrow \mathbb{R}$ be two 1D discrete signals. Their convolution can be written in the same form as \eqnref{graph_convolution_spatial}:
\begin{equation*}
	(f \ast g) [i] = \sum_{j=-\infty}^\infty f[j] g[i-j] = \langle T_i g,  f \rangle,
\end{equation*}
where $T_i g[j] = g[i-j]$ is, up to a flip, a translation operator. 
% nati: I think now the indexing is clear and will not confuse with the rest of the paper.
Similarly as \eqnref{graph_convolution_spatial}, the convolution of the signal $f$ by a kernel $g$ is the scalar product of $f$ with translated versions $T_i g$ of the kernel $g$.
% Signals and kernels are the same thing in classical signal processing precisely because there exists a proper definition of translation.
% For Michael: I know what you mean here, but it is slightly more complicated than that. 
Additionally, it turns out that the localization operator $\T_i h$ is a generalization of the translation operator on graphs. In the particular case where the Laplacian matrix $\L$ is circulant, $\T_i h$ is a translated version of $\T_j h$ for all $i, j$ and both convolutions are equivalent. We refer the reader to \citet[Sec 2.2]{perraudin2017stationary} for a detailed discussion of the connection between translation $T_i$ and localization $\T_i$.

@@

Following \citet{defferrard2016convolutional}, we define our filters as Chebyshev polynomials such as the convolution operation becomes
\begin{equation} \label{eqn:graph_convolution_cheby}
	h_\theta\left(\tL\right) \f = \U \left(\sum_{k=0}^K \theta_k T_k\left(\tilde{\bLambda}\right) \right) \U\trans \f = \sum_{k=0}^K \theta_k T_k\left(\tL\right) \f,
\end{equation}
where $\tL = \frac{2}{\lambda_{max}} \L - \b{I} = -\frac{2}{\lambda_{max}} \D^{-1/2} \W \D^{-1/2}$ is the rescaled Laplacian with eigenvalues $\tilde{\b \Lambda}$ in $[-1, 1]$ and $T_k(\cdot)$ is the Chebyshev polynomial of degree $k$, defined by the recursive relation $T_k(\tL) = 2\tL T_{k-1}(\tL) - T_{k-2}(\tL)$, $T_1(\tL) = \tL$, $T_0(\tL) = \b{I}$. While definitions \eqnref{graph_convolution_monomial} and \eqnref{graph_convolution_cheby} both allow the representation of the same set of filters, Chebyshev polynomials are preferred as they form an orthogonal basis on the interval $[-1, 1]$. Moreover, they factor out $\tL^{k-2}$ from $T_k(\tL)$, e.g., the 3-neighborhood is not mixed with the 5-neighborhood. That results in easier learning when optimizing the parameters $\theta$.
Finally, the coefficients can be computed such as the truncated Chebyshev expansions gives an approximate minimax polynomial of a given filter \citep{hammond2011wavelets}. While that is irrelevant when learning filters, it is useful when smoothing a signal on the sphere by filtering it with an approximated low-pass filter.
In practice, we found out in our experiment that Chebyshev polynomials are slightly more stable than the monomials used in~\cite{khasanova2017graphomni}(i.e. optimizing directly the $\theta_k$ in \eqnref{graph_convolution_monomial}).
We believe that this is due to their orthogonality (as explained above) and uniformity\footnote{The amplitude of Chebyshev polynomials is more or less constant over the domain $[-1, 1]$ independently of the order $k$. On the contrary the amplitude of $x^k$ is very different for $|x|\approx 0$ and $|x| \approx 1$.}.
% \nati{I added a sentence and I think we should not give much more details here. If we do an experiment, we can comment in the experiment section.}
% \todo{(i) scale of $\theta_k$: they are mostly uniform for Chebyshev polynomials, end exploding for monomials. Though SGD implicitly imposes a uniform prior on the learned parameters.}
% \todo{(ii) numerical stability of $\L^k \f$ v.s. $T_k(\L) \f$}
% \todo{(iii) in the code, initialize $\theta_k$ with stdev adapted to $F_{in}$ and $K$}

@@

% The advantage of schemes based on a base polyhedron is that cells are equal-area. While that is not required for the convolution operation when modeling with graphs (it can be corrected by setting edge weights appropriately, see \citet{khasanova2017graphomni})
% nope: can also do a regular pooling on the equirectangular. It preserves the sampling, and the convolution is adapted to the sampling thanks to the edge weights.

@@

Following \citet{khasanova2017tigranet}, we implement an optional statistical layer between the last graph convolutional layer and the first fully connected layer for the neural network.
This introduces the invariance of the output to isometric transformations, such as rotation of maps on the sphere.
The role of the layer is to compute global and location independent statistics from feature maps.
Besides providing invariance, the layer drastically compresses the representation of the data as the number of computed statistics $N_{stat}$ is usually much smaller than the size of the map $N_{pix}$. That compression results in much smaller fully connected layers, which usually contain a significant fraction of trainable parameters of a neural network.
The invariance and reduced representation should help the network to learn from fewer samples and to generalize better to unseen data. Yet another advantage is that a network built with such a statistical layer is agnostic to the size of the input data, i.e., it can classify maps which do not have the same number of pixels. Indeed, such global summarization is invariant to node permutation and is commonly used along graph convolutions to classify graphs of varying sizes \citep{duvenaud2015gcn, li2015gatedgnn}.

>>>>

% \documentclass[preprint,12pt,authoryear]{elsarticle}
% \documentclass[final,twocolumn,3p,times,authoryear]{elsarticle}
% \documentclass[final,twocolumn,3p]{elsarticle}
\documentclass[final,twocolumn,3p,times,sort&compress]{elsarticle}
%% Use the option review to obtain double line spacing
%% \documentclass[authoryear,preprint,review,12pt]{elsarticle}

@@

% TK: This looks OK, but the math is a bit ugly
\usepackage[T1]{fontenc}
\usepackage{txfonts} % to remove when done with todo list

% TK: This looks nice but math is spilling out
% \usepackage{lmodern}
% \usepackage[T1]{fontenc}

% TK: No difference
% \usepackage{helvet}

\usepackage[utf8]{inputenc}

@@

%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
\usepackage{color}
\usepackage{url}
% \usepackage{hyperref}
\usepackage[draft]{hyperref}
\usepackage{graphicx,array} \graphicspath{{figures/}}
\usepackage{subcaption}
\usepackage{amsmath}
\usepackage{siunitx}
\usepackage{bm}
\usepackage{aasmacros}
\usepackage{enumitem} % to remove when done with todo list
 % to remove when done with todo list

%% The lineno packages adds line numbers. Start line numbering with
%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
%% for the whole article with \linenumbers.
\usepackage{lineno}
\linenumbers

@@

\author[SDSC]{Nathanaël Perraudin}
\author[EPFL]{Michaël Defferrard}
\author[ETHZ]{Tomasz Kacprzak}
\author[ETHZ]{Raphael Sgier(?)}
% \author{TBD}

@@

Convolutional Neural Networks (CNNs) are now a cornerstone of the Deep Learning toolbox and have led to many breakthroughs in Artificial Intelligence.
So far, these networks have mostly been developed for regular Euclidean domains such as those supporting images, audio, or video.
Because of their success, CNN-based methods are becoming increasingly popular in Cosmology.
Cosmological data often comes in a form of spherical maps, which makes the use of traditional CNNs more complicated.
The commonly used pixelzation scheme for spherical maps is Hierarchical Equal Area isoLatitude Pixelisation (HEALPix).
We present a spherical CNN for analysis of full and partial HEALPix maps, which we call HealpixNet.
We construct a spherical CNNs by representing the sphere as a graph.
Graphs are versatile data structures which can represent pairwise relationships between objects or act as a discrete representation of a continuous manifold.
Using the graph-based representation, we define many of the standard CNN operations, such as convolution and pooling, with filters restricted to being radial.
This way, HealpixNet is a special case of a graph CNN, tailored to the HEALPix sampling of the sphere.
% TBD if we include it
% This approach is very efficient computationally than using spherical harmonics to perform convolutions, an alternative that has been explored previously.
An advantage of this representation is that the convolution operations with this approach are very efficient, which enables the algorithm to run very fast.
We demonstrate the method on a classification problem of mass maps from two cosmological models and compare the performance of the CNN with that of two baseline classifiers, based on the power spectrum and pixel density histogram.
Our experimental result shows that the performance of HealpixNet to always be better or equal to both of these baselines.
For high noise levels and for data covering only a smaller fraction of the sphere, HealpixNet achieves typically 10\% better classification accuracy than those baselines.
Finally, we propose a visualization of filters learned by the HealpixNet from the mass maps as some analysis tool for the neural network.
Code and examples are available at \todo{\url{https://github.com/SwissDataScienceCenter/HealpixNet}}.

@@

\section*{TODO list}
\begin{itemize}[noitemsep,topsep=0pt,parsep=0pt,partopsep=0pt]
    \item (MD) distinction between CNN and FCN
    \item (MD) check paragraph on our chosen architecture in the Experiments sections
    \item (MD) check vertex interpretation in Efficient convolutions. Powers of L are weights of paths.
    \item (TK) check part sphere notebook and whole sphere
    \item (NP) Put Raphael in the dataset author.
    \item (Probably OK) Beware to not confuse reader with synonyms, e.g. pixelization and sampling.
    \item (Should be done) Check names: we use HealpixNet, HealpixNet (CNN variant), HealpixNet (FCN variant)
    \item (I think it is OK) (NP) Check the wording. Be sure that the convolution on SO(3) is clear: we want to say (a) is the one associated with the SO(3) rotation group (b) can be performed  in the spectral domain by a multiplication (c) is equivariant to rotation (d) is described well in Section 2 of \citep{kondor2018clebsch}.
    \item (Everyone should check this) Notation: write element $i$ of vector $y$ as $y_i$, $y(i)$, or $y[i]$? Be consistent. Same for matrices. Keep indices.
    \item Check links in footnotes.
    \item (done!) (NP) Change equation after (2) and check all kroneker
    \item (done!) (TK) check text figure 1
    \item (done!) (TK) Related work in cosmo. The only cosmological applications of spherical CNNs in the related work are about projection on the tangent plane \citep{schmelze2017cosmologicalmodel, fluri2018deep, gupta2018nongaussianinformation}. Aren't there applications in cosmology for the other (c.f. related work section) approaches? Most of them are illustrated in omnidirectional imagery. If none, say it and mention some alternatives, e.g., wavelets?
    \item (done!) (TK) check if papers which appear as arxiv were not published in a conference or journal
    \item (done!) (TK) read everything
    \item (fail - I checked few fonts here: \url{de.sharelatex.com/learn/Font_typefaces}, but none looked better) (TK) Font: do we have to use times (set in the documentclass)? I (MD) find the rendering of math a bit ugly.
    \item (done!) (TK) fix the bibliography
    \item (done!) (TK) add small description of maps\_downsampled\_64.npz in the whole\_sphere notebook. \nati{We need to talk about that.}
    \item (done!) (TK) Example applications for dense prediction (classification and regression) for cosmology (color text in 2.8)

@@

% \todo{@Tomek: can you explain why we care about parts of the sphere? (that is an advantage of the graph model over the spherical harmonics)}
% \todo{@Tomek: How is the data acquired? Do telescopes and satellites take measurements according to the HEALPix sampling? (it is also an advantage of our method to be agnostic to the sampling, which might be useful to researchers working with missing data or different models)}
% \nati{@Tomek: I think what Michael has in mind is that the graph approach is very versatile and extendable. It can be used with other samplings (even irregular ones) or only part of the sphere. This flexibility might be an advantage to mention. In particular, compared to~\citep{cohen2018sphericalcnn} it allows us to do only do the required computation when only a part of the sphere is considered.}

@@

% cohen2017convolutional (ICML workshop) and cohen2018sphericalcnn (ICLR best paper) are redundant
A first direction to generalize CNNs to spherical data was to apply a standard 2D CNN to a grid discretisation of the sphere~\citep{boomsma2017spherical,su2017sphericalconv,coors2018spherenet}.
While this approach uses the well-developed 2D convolution and hierarchical pooling, it is restricted to specific grid sampling schemes that are highly irregular, and hence inherently deform the domain where the data is supported.
A second direction attempted by~\citep{cohen2018sphericalcnn} was to leverage the spherical Fourier transform and to perform the convolution associated to the SO(3) rotation group in the spectral domain, thanks to the convolution theorem.
Contrarily to the first direction, the resulting convolution is rotational equivariant, meaning that a rotation of the input implies the same rotation of the output.
This is an important property as translation invariance is probably one of the main reasons why classic CNNs are successful in the first place.
Nevertheless, this approach is computationally expensive even if a fast spherical Fourier transforms is used.
%(They exist for particular samplings.)
Finally, the use of a graph to model the discretised sphere was considered by~\citep{khasanova2017graphomni}, towards the goal of designing a network that is invariant to rotations.
While its architecture accommodates these invariances on the sphere and remains fairly efficient, it did not take advantage of a hierarchical pixelisation and use a simple grid discretisation of the sphere.
Moreover, the statistical layer, which introduces the invariance, needs to be hand-crafted to capture statistics of interest.

This work is the first to design a spherical convolutional neural network for cosmological applications.
Previously, several other machine learning algorithms were extended to the sphere, often also including the third, radial direction.
Most notably, sparsity-based methods were developed, such as wavelets \citep{Leistedt2013flaglets,Leistedt2015spinwavelets}, ridgelets \citep{McEwen2015ridgelet} and curvelets \citep{Chan2017curvelets,Starck2009polarized}.

%\todo{better identify the shortcomings: (i) part of sphere / missing data, (ii) computational efficiency, (iii) equivarience}
% \nati{The part of the sphere/missing data is not really there, but I still do not know how to include it!}

We introduce an efficient spherical CNN formulation which uses (i) convolutions on graphs and (ii) hierarchical pooling, while working with the Hierarchical Equal Area isoLatitude Pixelisation (HEALPix) sampling.
The main idea is to represent the sphere $S^2$, a 2-dimensional manifold embedded in $\mathbb{R}^3$, as a graph of connected pixels.
The length (or weight) of the shortest path between two pixels is an approximation of the geodesic distance between them on the manifold.
The use of this approach is enabled by recent developments in graph signal processing, where fundamental tools, such as for example the Fourier transform, were defined for graphs \citep{shuman2013emerging}.
Our approach uses the graph CNN formulation introduced by \citep{defferrard2016convolutional}, which can accommodate any sampling of the sphere.
Using a $k$-nearest neighbours graph, the convolution operation costs $\mathcal{O}(N_{pix})$ operations, where $N_{pix}$ is the number of pixels.
This is the lowest possible complexity for a convolution operation without approximations (e.g., by sketching).
The flexibility of modeling the data domain with a graph allows to easily model data that spans only a part of the sphere, or data that is not uniformly sampled.
Our hierarchical pooling strategy exploits a hierarchical pixelisation of the manifold to analyse the data at multiple scales.
% Such hierarchical pixelisations of the sphere have been well studied.

@@

We give a practical demonstration of the application of our package to cosmological model discrimination using maps of projected mass distribution on the sky \citep{chang2017curvedsky}.
This kind of maps can be created using the gravitational lensing technique \citep[see][for review]{BartelmannSchneider2001weak}.
Our maps are similar to the ones used by \citep{schmelze2017cosmologicalmodel}.
In a simplified scenario, we classify convergence maps on parts of a sphere into two cosmological models.
These models were designed to have very similar angular power spectrum for the angular wavenumbers $\ell < 1000$.
We compare the performance of our spherical CNN to a baseline algorithm, an SVM classifier which takes pixel histograms or power spectrum densities (PSDs) of these maps as input.
The comparison is made as a function of the additive noise level and the area of the sphere used in the analysis.
% Results show that our model is better at discriminating the maps and is more resilient to noise.

@@

A CNN is composed of the following main building blocks~\citep{lecun1998cnn}:
(i) a convolution,
(ii) a non-linearity,
(iii) a down-sampling operation,
(iv) a pooling operation, and
(v), optionally,  normalization.\footnote{\citep{ioffe2015batchnorm} have shown that batch normalization helps training. We verified this experimentally in our setting as well.}
Our architecture is depicted in~\figref{architecture} and discussed in greater details in \secref{architecture}. As operations (ii) and (v) are point-wise, they do not depend on the data domain. The pooling operation is simply a permutation invariant aggregation function which does not need to be adapted either. The convolution and down-sampling operations, however, need to be generalized from Euclidean domains to the sphere.

@@

Before doing any numerical analysis on the sphere, one first has to choose a tessellation, i.e., an exhaustive partition of the sphere into finite area elements, where the data under study is quantized.
The simplicity of the spherical form belies the intricacy of global analysis on the sphere.
There is no known point set which achieves the analogue of uniform sampling in Euclidean space and allows exact and invertible discrete spherical harmonic decompositions of arbitrary but band-limited functions.
Any existing proposition of practical schemes for the discrete treatment of such functions on the sphere introduces some systematic error dependent on the global properties of the point set.
While our method is applicable to any pixelisation of the sphere, two details depend on the chosen sampling: (i) the choice of neighbours in the construction of the graph, and (ii) the choice of parent nodes when coarsening the graph.
As HEALPix~\citep{gorski2005healpix} is the most popular sampling used in cosmology, our target application, we tailor the method to that particular sampling in the subsequent exposition.
%We'll note in later sections where the method should be adapted for a different sampling.

@@

% HEALPix is an acronym for Hierarchical Equal Area isoLatitude Pixelization of a sphere.
The HEALPix pixelisation produces a hierarchical subdivision of a spherical surface where each pixel covers the same surface area as every other pixel.
A hierarchy is desired for the data locality in the computer memory.
Equal area is advantageous because white noise generated by the signal receiver gets integrated exactly into white noise in the pixel space.
Isolatitude is essential for the implementation of a fast spherical transform.
HEALPix is the sole pixelisation scheme which satisfies those three properties.

@@

Our graph is constructed as an approximation of the continuous 2D manifold that is the surface of the sphere.
Indeed, \citep{belkin2007convergence} showed that the graph Laplacian converges to the Laplace-Beltrami when the number of pixels goes to infinity.
While our construction will not exactly respect the setting defined by \citep{belkin2007convergence}, we observe empirically strong evidence of convergence.

@@

Finally, the graph convolution can be interpreted in the vertex domain as a scalar product with localizations $\T_i h$ of the kernel $h$. Indeed, the result of the convolution of the signal $\f$ with the kernel $h$ is
\begin{equation} \label{eqn:graph_convolution_spatial}
	(h(\L) \f)_i = \langle \T_i h(\L), \f \rangle = \langle h(\L) \b \delta^i, \f \rangle.
\end{equation}
To make the parallel with the classical 1D convolution, let $f, g\in \mathbb{Z} \rightarrow \mathbb{R}$ be two 1D discrete signals. Their convolution can be written in the same form as \eqnref{graph_convolution_spatial}:
\begin{equation*}
	(f \ast g) [i] = \sum_{j=-\infty}^\infty f[j] g[i-j] = \langle T_i g,  f \rangle,
\end{equation*}
where $T_i g[j] = g[i-j]$ is, up to a flip, a translation operator.
% nati: I think now the indexing is clear and will not confuse with the rest of the paper.
Similarly as \eqnref{graph_convolution_spatial}, the convolution of the signal $f$ by a kernel $g$ is the scalar product of $f$ with translated versions $T_i g$ of the kernel $g$.
% Signals and kernels are the same thing in classical signal processing precisely because there exists a proper definition of translation.
% For Michael: I know what you mean here, but it is slightly more complicated than that.
Additionally, it turns out that the localization operator $\T_i h$ is a generalization of the translation operator on graphs. In the particular case where the Laplacian matrix $\L$ is circulant, $\T_i h$ is a translated version of $\T_j h$ for all $i, j$ and both convolutions are equivalent. We refer the reader to \citep[Sec 2.2]{perraudin2017stationary} for a detailed discussion of the connection between translation $T_i$ and localization $\T_i$.

@@

Following \citep{defferrard2016convolutional}, we define our filters as Chebyshev polynomials such as the convolution operation becomes
\begin{equation} \label{eqn:graph_convolution_cheby}
	h_\theta\left(\tL\right) \f = \U \left(\sum_{k=0}^K \theta_k T_k\left(\tilde{\bLambda}\right) \right) \U\trans \f = \sum_{k=0}^K \theta_k T_k\left(\tL\right) \f,
\end{equation}
where $\tL = \frac{2}{\lambda_{max}} \L - \b{I} = -\frac{2}{\lambda_{max}} \D^{-1/2} \W \D^{-1/2}$ is the rescaled Laplacian with eigenvalues $\tilde{\b \Lambda}$ in $[-1, 1]$ and $T_k(\cdot)$ is the Chebyshev polynomial of degree $k$, defined by the recursive relation $T_k(\tL) = 2\tL T_{k-1}(\tL) - T_{k-2}(\tL)$, $T_1(\tL) = \tL$, $T_0(\tL) = \b{I}$. While definitions \eqnref{graph_convolution_monomial} and \eqnref{graph_convolution_cheby} both allow the representation of the same set of filters, Chebyshev polynomials are preferred as they form an orthogonal basis on the interval $[-1, 1]$. Moreover, they factor out $\tL^{k-2}$ from $T_k(\tL)$, e.g., the 3-neighborhood is not mixed with the 5-neighborhood. That results in easier learning when optimizing the parameters $\theta$.
Finally, the coefficients can be computed such as the truncated Chebyshev expansions gives an approximate minimax polynomial of a given filter \citep{hammond2011wavelets}. While that is irrelevant when learning filters, it is useful when smoothing a signal on the sphere by filtering it with an approximated low-pass filter.
In practice, we found out in our experiment that Chebyshev polynomials are slightly more stable than the monomials used in~\cite{khasanova2017graphomni} (i.e. optimizing directly the $\theta_k$ in \eqnref{graph_convolution_monomial}).
We believe that this is due to their orthogonality (as explained above) and uniformity\footnote{The amplitude of Chebyshev polynomials is more or less constant over the domain $[-1, 1]$ independently of the order $k$. On the contrary the amplitude of $x^k$ is very different for $|x|\approx 0$ and $|x| \approx 1$.}.
% \nati{I added a sentence and I think we should not give much more details here. If we do an experiment, we can comment in the experiment section.}
% \todo{(i) scale of $\theta_k$: they are mostly uniform for Chebyshev polynomials, end exploding for monomials. Though SGD implicitly imposes a uniform prior on the learned parameters.}
% \todo{(ii) numerical stability of $\L^k \f$ v.s. $T_k(\L) \f$}
% \todo{(iii) in the code, initialize $\theta_k$ with stdev adapted to $F_{in}$ and $K$}

@@

% The advantage of schemes based on a base polyhedron is that cells are equal-area. While that is not required for the convolution operation when modeling with graphs (it can be corrected by setting edge weights appropriately, see \citep{khasanova2017graphomni})
% nope: can also do a regular pooling on the equirectangular. It preserves the sampling, and the convolution is adapted to the sampling thanks to the edge weights.

@@

Following \citep{khasanova2017tigranet}, we implement an optional statistical layer between the last graph convolutional layer and the first fully connected layer for the neural network.
This introduces the invariance of the output to isometric transformations, such as rotation of maps on the sphere.
The role of the layer is to compute global and location independent statistics from feature maps.
Besides providing invariance, the layer drastically compresses the representation of the data as the number of computed statistics $N_{stat}$ is usually much smaller than the size of the map $N_{pix}$. That compression results in much smaller fully connected layers, which usually contain a significant fraction of trainable parameters of a neural network.
The invariance and reduced representation should help the network to learn from fewer samples and to generalize better to unseen data. Yet another advantage is that a network built with such a statistical layer is agnostic to the size of the input data, i.e., it can classify maps which do not have the same number of pixels. Indeed, such global summarization is invariant to node permutation and is commonly used along graph convolutions to classify graphs of varying sizes \citep{duvenaud2015gcn, li2015gatedgnn}.
